
# Functional programming


## Introduction

- Using functions is an easy way to quickly improve reproducibility - no more copy/paste errors!
- The concept is "do not repeat yourself"


### Function usage:

Example: You have a typo that repeats in all of your datasets (for example "a", instead of "A" and "c" instead of "C")
```{r, eval = F}
library(tidyverse)

original_data1 = tibble(letters = c("a", "B", "c","D"),
                        numbers = c(1,22,3,4))

original_data2 = tibble(letters = c("c","a", "D", "B"),
                        numbers = c(3,4,1,22))

original_data3 = tibble(letters = c("D","c", "B", "a"),
                        numbers = c(22,3,4,1))
```

Doing it manually
```{r}
new_data1 = original_data1 %>% 
  mutate(letters = recode(letters,"a" = "A"),
         numbers = recode(numbers,"22" = 2))

new_data2 = original_data2 %>% 
  mutate(letters = recode(letters,"a" = "A"),
         numbers = recode(numbers,"22" = 2))

new_data3 = original_data3 %>% 
  mutate(letters = recode(letters,"a" = "A"),
         numbers = recode(numbers,"22" = 2))
```


Doing it with a function
```{r}

fix_typo = function(data){
  
  temp_data = data %>% 
    mutate(letters = recode(letters,"a" = "A"),
         numbers = recode(numbers,"22" = 2))
  
  return(temp_data)
}

new_data1 = fix_typo(original_data1);new_data1
new_data2 = fix_typo(original_data2)
new_data3 = fix_typo(original_data3)

```

### Function parts:

function_name = function(replace_data, re){

  temporary_data = replace_data %>%
    actions...

  return(temporary_data)

}




### The state of your program

1.Create a function that prints a sentence:
```{r}
printer = function(name, food){
  print(paste0(name, " likes ", food))
}

printer("Gabe", "pizza")
```
2. List all objects you have:
```{r, eval = F}
ls()
```
- Note that the function is listed, but no new objects were created.

3. Embed the food options within the function and print a new item each time.
```{r}
printer = function(name){
  food = sample(c("pizza", "falafel", "cake"), 1)
  print(paste0(name, " likes ", food))
}

printer("Gabe")
```

- You can set seed to stop generating random draws
```{r}
printer = function(name, seed){
  set.seed(seed)
  food = sample(c("pizza", "falafel", "cake"), 1)
  print(paste0(name, " likes ", food))
}

printer("Gabe", seed = 123124)
```
4. Create an object with the randomly selected food:
```{r}
printer = function(name){
  food = sample(c("pizza", "falafel", "cake"), 1)
  
  food_vector <<- append(vector(), food)
  
  print(paste0(name, " likes ", food))
   }

printer("Gabe")
```
```{r}
ls()
```
```{r}
food_vector
```

5. Create a list with all of the the randomly selected food items:
```{r}
printer = function(name){
  food = sample(c("pizza", "falafel", "cake"), 1)
  
  if(exists("food_list")){
    food_list <<- append(food_list, food)
  } else {
    food_list <<- append(list(), food)
  }
  
  print(paste0(name, " likes ", food))
   }

printer("Gabe")
```
```{r}
food_list
```

However, we want to avoid using functions that change the state as much as
possible!

```{r, eval = F}
printer = function(name, food_list = list()){
  
  food = sample(c("pizza", "falafel", "cake"), 1)
  
  food_list = append(food_list, food)
  
  print(paste0(name, " likes ", food))
   }

printer("Gabe")


```

The difference now is that we made `food_list` the second argument of the
function. Also, we defined it as being optional by writing:

```{r, eval = F}
food_list = list()
```


## Writing good functions

### 1) Create self-contained functions

That is, all external arguments are listed as function parameters. 

```{r, eval = F}
bad = function(x){
  x + y
}
```

```{r, eval = F}
good <- function(x, y){
  x + y
}
```

PS: It's possible to define a function that explicitly takes another function
as an input:

```{r}
calc <- function(number, other_function){
  other_function(number)
}
```

```{r}
calc(4, sqrt)

calc(10, log10)
```

If you don't know how many arguments the function you're wrapping has, you can use the `...`:

```{r}
calc <- function(number, other_function, ...){
  other_function(number, ...)
}
```

```{r}
data = c(1, 2, NA, 3) 

calc(data, mean, na.rm = TRUE)

```


### 2) Create error warnings

Prevent errors to occurr unnoticed by creating stop/warning messages. 

For example, some functions cannot be used with negative values:
```{r}
data = sqrt(-5)
```
```{r}
log10(-10)
```

It is useful to redefine these functions to raise an error function instead:

```{r}
strict_sqrt <- function(x){

  if(x < 0) stop("x is negative")

  sqrt(x)

}
```

This function now throws an error for negative `x`:

```{r, eval = F}
strict_sqrt(-10)
```

Functions like this (that return functions) are called *function factories*.

The `{purrr}` package also comes with function factories that you might find
useful (`{possibly}`, `{safely}` and `{quietly}`).


### Optional arguments

It is possible to make functions’ arguments optional, by using `NULL`. 

```{r}
sum_y <- function(x, y = NULL){

  if(is.null(y)){
    print("optional argument y is NULL")
    
    x
  } 
  
  else {
    print("y is present")
    
    x+y
  }

}

```

```{r}
x = c(1,2,3,4)
y = c(4,3,2,1)

sum_y(x)
```
```{r}
sum_y(x, y)
```

### Recursive functions

A function that calls itself in its own body is called a recursive function. 

In R, they are quite slow. 

```{r}
fact_iter <- function(n){
  result = 1
  for(i in 1:n){
    result = result * i
  }
  result
}


fact_recur <- function(n){
  if(n == 0 || n == 1){
  result = 1
  } else {
    n * fact_recur(n-1)
  }
}
```

Using the `{microbenchmark}` package we can benchmark the code:

```{r, eval = FALSE}
microbenchmark::microbenchmark(
  fact_recur(50),
  fact_iter(50)
)
```

We see that the recursive factorial function is 10 times slower than the iterative version. If you're working with more complex functions, this is a problem. 

Make recursive functions faster using *trampolining* with such packages: 

[`{trampoline}`](https://rdinnager.github.io/trampoline/)^[https://rdinnager.github.io/trampoline/]

 [`{memoise}`](https://memoise.r-lib.org/)^[https://memoise.r-lib.org/] 



### Anonymous functions

It is possible to define a function and not give it a name. 

```{r, eval = F}
function(x)(x+1)(10)
```

Or:

```{r, eval = F}
(\(x)(x+1))(10)
```


### Function lists

- Write simple functions that only perform one task. 
- They are easier to maintain, test, document and debug. 
- Then chain the small functions using the `|>` operator.

```
a |> f() |> g() |> h()
```
where `a` is for example a path to a data set, and where `f()`, `g()` and `h()` successively read, clean, and plot the data.


- Lists are the second important ingredient of functional programming. 
- They are extremely flexible, and most of the very complex objects classes that you manipulate are actually lists, but just fancier. For example, a data frame is a list:

```{r}
data(mtcars)

typeof(mtcars)
```

A fitted model is a list:

```{r}
my_model <- lm(hp ~ mpg, data = mtcars)

typeof(my_model)
```
A `ggplot` is a list:

```{r}
library(ggplot2)

my_plot <- ggplot(data = mtcars) +
  geom_line(aes(y = hp, x = mpg))

typeof(my_plot)
```

### Lists can hold many things

If you write a function that needs to return many objects, the only solution is to place them inside a list. 


```{r}
cars = function(data){
my_summary = mean(data$hp)
my_model = lm(hp ~ mpg, data = mtcars)
my_plot = ggplot(data = mtcars) +
  geom_line(aes(y = hp, x = mpg))

result = list(my_summary = my_summary, my_model = my_model, my_plot= my_plot)

return(result)
  
}
```


```{r}
res = cars(mtcars)
```


```{r}
res$my_summary
```


```{r}
res$my_model
```


```{r}
res$my_plot

```

### Lists as the cure to loops

Loops are incredibly useful, and you are likely familiar with them. The problem with loops is that they are a concept from iterative programming, not functional programming, and this is a problem because loops rely on changing the state of your program to run. For example, let's suppose that you wish to use a for-loop
to compute the sum of the first 100 integers:

```{r}
result <- 0

for (i in 1:100){
  result <- result + i
}

print(result)
```

If you run `ls()` now, you should see that there's a variable `i` in your global environment. This could cause issues further down in your pipeline if you need to re-use `i`. 


```{r}
my_list = list(
  c(5, 8, 2, 9), 
  'cat', 
  'dog', 
  c('koala', 'panda', 'rabbit'), 
  TRUE, 
  3.14)
```

Define a function to calculate the length of each element

```{r}
length_fun <- function(x) {
  return(length(x))
}
```

Use lapply to apply the length_fun function to each element of the list
```{r}
lapply(my_list, length_fun)
```

Alternatively, if you have a single data set, you can apply: 

```{r}
my_data = tibble(col1 = c(1,2,3,4),
              col2 = c(5,6,7,8),
              col3 = c(9,10,11,12)) 


# create a function
midrange = function(x){(min(x) + max(x)) / 2}
```


```{r}
# Apply the midrange function to each row of the iris dataset
apply(my_data, 1, midrange)
```


```{r}
# Apply the midrange function to each column of the dataset
apply(my_data, 2, midrange)

```

#### `Filter()`. 
`Filter()` filters data where the elements of objects within the list satisfy a predicate.

```{r}
a_list = list(seq(1, 5),
         "Hey",
         c(8942387,12323))

Filter(is.character, a_list)
```
#### `Negate()` 

 `Negate()` is a function factory that takes a boolean function as an input and returns the opposite boolean function. 
 
 

```{r}
Filter(Negate(is.character),a_list)
```

#### `local()` 

 `local()` runs code in a temporary environment that gets discarded at the end, leaving the state of the program intact.

```{r}
local({a = 2})
```

Variable `a` was created inside this local environment. Checking if it exists now
yields `FALSE`:

```{r}
ls()
```



### purrr

`apply()` is a family of functions that loop over a list and applying a function to the elements of the list
`lapply()` 
`sapply()`
`vapply()`
`mapply()`
`tapply()`

but it is not quite easy to remember which function does what, and there are inconsistencies in the order of arguments and types of output.


`{purrr}` solves this issue by offering the `map()` family of functions, which behave in a very consistent way. 

see functions `map`, `walk`, `reduce()`,`accumulate()`

Read through the [documentation of the
package](https://purrr.tidyverse.org/reference/index.html)^[https://purrr.tidyverse.org/reference/index.html] and take the time to learn about all it has to offer.

### withr

`{withr}` has many useful functions which I encourage
you to [familiarize yourself with](https://withr.r-lib.org/reference/index.html)^[https://withr.r-lib.org/reference/index.html].



### Functional OOP 

Important topic if you want to program packages, but not addressed in details in this chapter.

You can create functions that will do different things depending the type of object the user gives.

For example

```{r}
print(5)
print(head(mtcars))
print(str(mtcars))
```

There are actually many `print()` functions. For example, type
`print.data.frame` is the `print` function for `data.frame` objects. 

So what `print()` does, is look at the class of its argument `x`, and then look for the right `print` function to call. 

To learn more about R’s different OOP systems, take a look at @wickham2019.

## Closing remarks:

- Write pure self-contained functions
- Replace loops with higher-order functions (`lapply()`,`map()`,...) 



_______________________________________________________________

If there is time: create rooms to work together on "Data frames"


### Data frames
As mentioned in the introduction of this section, data frames are a special type
of list of atomic vectors. This means that just as I can use `lapply()` to
compute the square root of the elements of an atomic vector, as shown
previously, I can also operate on all the columns of a data frame. For example,
it is possible to determine the class of every column of a data frame like this:

```{r}
lapply(iris, class)
```

Unlike a list however, the elements of a data frame must be of the same length.
Data frames remain very flexible though, and using what we have learned until
now it is possible to use the data frame as a structure for all our
computations. For example, suppose that we have a data frame that contains data
on unemployment for the different subnational divisions of the Grand-Duchy of
Luxembourg, the country the author of this book hails from. Let’s suppose that I
want to generate several plots, per subnational division and per year.
Typically, we would use a loop for this, but we can use what we’ve learned here,
as well as some functions from the `{dplyr}`, `{purrr}`, `{ggplot2}` and
`{tidyr}` packages. I will be downloading data that I made available inside a
package, but instead of installing the package, I will download the `.rda` file
directly (which is the file format of packaged data) and then load that data
into our R session (instead of downloading from the long Github url, I download
the data from a shortened *is.gd* link):

```{r}
# Create a temporary file
unemp_path <- tempfile(fileext = ".rda")

# Download the data and save it to the path of the temporary file
# avoids having to install the package from Github
download.file(
  "https://is.gd/l57cNX",
  destfile = unemp_path)

# Load the data. The data is now available as 'unemp'
load(unemp_path)

```

Let’s load the required packages and take a look at the data:

```{r}
library(dplyr)
library(purrr)
library(ggplot2)
library(tidyr)

glimpse(unemp)

```

Column names are self-descriptive, but the `level` column needs some
explanations. `level` contains the administrative divisions of the country, so
the country of Luxembourg, then the Cantons and then the Communes.

Remember that Luxembourg can refer to the country, the canton or the commune of
Luxembourg. Now let’s suppose that I want a separate plot for the three communes
of Luxembourg, Esch-sur-Alzette and Wiltz. Instead of creating three separate
data frames and feeding them to the same ggplot code, I can instead take
advantage of the fact that data frames are lists, and are thus quite flexible.
Let’s start with filtering:

```{r}
filtered_unemp <- unemp %>%
  filter(
    level == "Commune",
    place_name %in% c("Luxembourg", "Esch-sur-Alzette", "Wiltz")
   )

glimpse(filtered_unemp)
```

We are now going to use the fact that data frames are lists, and that lists can
hold any type of object. For example, remember this list from before where one
of the elements is a data frame, and the second one a formula:

```{r}
list(
  "a" = head(mtcars),
  "b" = ~lm(y ~ x)
  )
```

`{dplyr}` comes with a function called `group_nest()` which groups the data
frame by a variable (such that the next computations will be performed
group-wise) and then nests the other columns into a smaller data frame. Let’s
try it and see what happens:


```{r}
nested_unemp <- filtered_unemp %>%
  group_nest(place_name)
```

Let’s see what this looks like:

```{r}
nested_unemp
```

`nested_unemp` is a new data frame of 3 rows, one per commune
("Esch-sur-Alzette", "Luxembourg", "Wiltz"), and of two columns, one for the
names of the communes, and the other contains every other variable inside a
smaller data frame. So this is a data frame that has one column where each
element of that column is itself a data frame. Such a column is called a
list-column. This is essentially a list of lists.

Let’s now think about this for a moment. If the column titled `data` is a list
of data frames, it should be possible to use a function like `map()` or
`lapply()` to apply a function on each of these data frames. Remember that
`map()` or `lapply()` require a list of elements of whatever type and a function
that accepts objects of this type as input. So this means that we could apply a
function that plots the data to each element of the column titled `data`. Since
each element of this column is a data frame, this function needs a data frame as
an input. As a first and simple example to illustrate this, let’s suppose that we
want to determine the number of rows of each data frame. This is how we would do
it:

```{r}
nested_unemp %>%
  mutate(nrows = map(data, nrow))
 # ’data’ is the name of
 # the list-column that contains
 # the smaller data frames
```

The new column, titled `nrows` is a list of integers. We can simplify it by
converting it directly to an atomic vector of integers by using `map_int()`
instead of `map()`:

```{r}
nested_unemp %>%
  mutate(nrows = map_int(data, nrow))
```

Let’s try a more complex example now. What if we want to filter rows (of course,
the simplest way would be to filter the rows we need before nesting the
data frame)? We need to apply the function `filter()` where its first argument
is a data frame and the second argument is a predicate:

```{r}
nested_unemp %>%
  mutate(nrows = map(data, \(x)filter(x, year == 2015)))
```

In this case, we need to use an anonymous function. This is because `filter()`
has two arguments and we need to make clear what it is we are mapping over and
what argument stays fixed; we are mapping over (iterating) the data frames but
the predicate `year == 2015` stays fixed.

We are now ready to plot our data. The best way to continue is to first get the
function right by creating one plot for one single commune. Let’s select the
dataset for the commune of `Luxembourg`:

```{r}
lux_data <- nested_unemp %>%
  filter(place_name == "Luxembourg") %>%
  unnest(data)
```

To plot this data, we can now write the required `ggplot2()` code:

```{r}
ggplot(data = lux_data) +
  theme_minimal() +
  geom_line(
    aes(year, unemployment_rate_in_percent, group = 1)
   ) +
  labs(title = "Unemployment in Luxembourg")

```

To turn the lines of code above into a function, you need to think about how
many arguments that function would have. There is an obvious one, the data
itself (in the snippet above, the data is the `lux_data` object). Another one
that is less obvious is in the title:

```
labs(title = "Unemployment in Luxembourg")
```

Ideally, we would want that title to change depending on the data set. So we
could write the function like so:

```{r}
make_plot <- function(x, y){
  ggplot(data = x) +
    theme_minimal() +
    geom_line(
      aes(year, unemployment_rate_in_percent, group = 1)
      ) +
    labs(title = paste("Unemployment in", y))
}

```

Let’s try it on our data:

```{r}
make_plot(lux_data, "Luxembourg")
```

Ok, so now, we simply need to apply this function to our nested data frame:

```{r}
nested_unemp <- nested_unemp %>%
  mutate(plots = map2(
    .x = data, # column of data frames
    .y = place_name, # column of commune names
    .f = make_plot
  ))

nested_unemp
```

If you look at the `plots` column, you see that it is a list of `gg` objects:
these are our plots. Let’s take a look at them:

```{r}
nested_unemp$plots
```

We could also have used an anonymous function (but it is more difficult to get
right):

```{r}
nested_unemp %>%
  mutate(plots2 = map2(
    .x = data,
    .y = place_name,
    .f = \(.x,.y)(
                ggplot(data = .x) +
                  theme_minimal() +
                  geom_line(
                    aes(year, unemployment_rate_in_percent, group = 1)
                   ) +
                  labs(title = paste("Unemployment in", .y))
                  )
           )
         ) %>%
  pull(plots2)

```

This list-column based workflow is extremely powerful and I highly advise you to
take the required time to master it. Remember, we never want to have to repeat
ourselves. This approach might seem more complicated than calling `make_plot()`
three times, but imagine that you need to do this for several countries, several
variables, etc... What are you going to do, copy and paste code everywhere? This
gets very tedious and more importantly, very error-prone, because now you’ve
just introduced many points of failure by having so much copy-pasted code. You
could of course use a loop instead of this list-column based workflow. But as
mentioned, the issue with loops is that you have to interact with the global
environment, which can lead to other issues. But whatever you end up using, you
need to avoid copy and pasting at all costs.


