---
title: "R-Ladies Bergen Kick-off Meeting"
author: "Julia, Jonelle & Astrid"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    logo: R-Ladies-horiz-bergen.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library( ggplot2 )
library( tidyverse )
```

# R-Ladies Global

## What is R-Ladies?

- A global community of ladies who use R
- that seeks to _support_ and _inspire_ others
- through **teaching**, **sharing**, and **mentorship**

# R-Ladies Bergen

## Who are we?

- Jonelle Dickow Villar
- Julia Romanowska
- Astrid Lunde

## Meetup group

- 45 members already!
- who are we?
- why did we gather here?

# Poll results

## How advanced in R are you?

```{r poll1-data,echo=FALSE}
# updated by Julia on 25 Nov.2019
subtitle <- "(out of 45 registered on Meetup)"

# here - create the tibble
poll.q1 <- tribble( ~q.alt, ~answers,
                    "I want to learn!", 3,
                    "I’ve learned basics and I want more!", 9,
                    "R is one of many programming languages I know", 3,
                    "I am Hadley Wickham (she/her) ;)", 1
                  )
# explicitly setting the order of labels
poll.q1$q.alt <- factor( poll.q1$q.alt, levels = rev( poll.q1$q.alt ), ordered = TRUE )
tot.ans.q1 <- sum( poll.q1$answers )

y.ticks <- seq( 0, 10, 2 )
ggplot( poll.q1, aes( q.alt, answers, fill = q.alt ) ) +
  geom_col() + coord_flip() +
  theme_minimal() +
  scale_y_continuous( breaks = y.ticks, labels = as.character( y.ticks ) ) +
  theme( 
    plot.title = element_text( hjust = 1 ),
    plot.subtitle = element_text( hjust = 1 ),
    legend.position = "none"
  ) +
  ylab("") + xlab("") +
  labs( title = paste( "Total of", tot.ans.q1, "R-Ladies answered"),
        subtitle = subtitle )
```

## What is the field where you (would) use R the most?

```{r poll2-data,echo=FALSE}
poll.q2 <- tribble( ~q.alt, ~answers,
    "Bioinformatics / structural bioinformatics", 3,
    "Statistics", 7,
    "Applied mathematics", 0,
    "Political science", 0,
    "Economics", 2,
    "Visualization", 0,
    "History / Literature", 0,
    "Meteorology", 0,
    "Cheminformatics", 0,
    "Biology", 2,
    "AI / Machine learning", 0,
    "Other", 3
)
poll.q2$q.alt <- factor( poll.q2$q.alt, levels = rev( poll.q2$q.alt ), ordered = TRUE )
tot.ans.q2 <- sum( poll.q2$answers )

ggplot( poll.q2, aes( q.alt, answers, fill = q.alt ) ) +
  geom_col() + coord_flip() +
  theme_minimal() +
  theme( 
    plot.title = element_text( hjust = 1 ),
    plot.subtitle = element_text( hjust = 1 ),
    legend.position = "none"
  ) +
  ylab("") + xlab("") +
  labs( title = paste( "Total of", tot.ans.q2, "R-Ladies answered"),
        subtitle = subtitle )
```

## What do you use R for?

```{r poll3-data,echo=FALSE}
poll.q3 <- tribble( ~q.alt, ~answers,
    "Plotting", 1,
    "Organizing tables", 0,
    "Exploring the data", 0,
    "Specific analyses", 0,
    "Everything", 12,
    "Nothing (yet!)", 3
)
poll.q3$q.alt <- factor( poll.q3$q.alt, levels = rev( poll.q3$q.alt ), ordered = TRUE )
tot.ans.q3 <- sum( poll.q3$answers )

y.ticks <- seq( 0, 12, 2 )
ggplot( poll.q3, aes( q.alt, answers, fill = q.alt ) ) +
  geom_col() + coord_flip() +
  theme_minimal() +
  scale_y_continuous( breaks = y.ticks, labels = as.character( y.ticks ) ) +
  theme( 
    plot.title = element_text( hjust = 1 ),
    plot.subtitle = element_text( hjust = 1 ),
    legend.position = "none"
  ) +
  ylab("") + xlab("") +
  labs( title = paste( "Total of", tot.ans.q3, "R-Ladies answered"),
        subtitle = subtitle )
```

# Aim of the meeting

## Aims

- Get to know one another
- Learn something new in R
- Get ideas for future meetings
- Introduce R-Ladies Global on social media 

## Tips for future meetings 

- **Themed evenings**
  * GitHub, Tidyverse, R Markdown, data visualization, functional programming, machine learning 

- **SatRday workshops** - mentoring hands-on 

- **Lightening talks** presenting research projects 

- **Hackathon-events** such as Reproducibility Hackathon for
  * teaching tools for sharing reproducible code
  * gaining experience with reproducing `#opendata` `#opencode` papers
  * connecting with `#openscience` fans from other disciplines

## Warm up!

- Groups of 4 or 5

- Say hi to everyone within the group
  - why did they come here?
  - where do they use R?

- Discuss: *how do you want future meetings to look like?*

##  Check out R-Ladies on social media 

- Join the directory:  http://rladies.org/directory/

- **Twitter**

  - R-Ladies Bergen:  @RLadiesBergen

  - R-Ladies Global:  @RLadiesGlobal

  - We are R-Ladies:  @WeAreRLadies

- **Facebook:**           R-Ladies-Bergen

- Sign up on **Meetup**:  https://meetup.com/rladies-bergen/

- **Slack** workspace: `rladies-norway`

- **github**: https://github.com/rladies/meetup-presentations_bergen

## Save the date!

The European R users meeting (eRum) comes to Milan.

The event will feature multiple parallel sessions on topics

- data science, 
- statistics
- machine learning

With international speakers, lightening talks and workshops.  

27-30 May, 2020 

R-Ladies from different European chapters will be there!  

http://2020.erum.io
