---
title: "R-Ladies Bergen Kick-off Meeting"
author: "Julia & Jonelle"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    logo: R-Ladies-horiz-bergen.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library( ggplot2 )
library( tidyverse )
```

# R-Ladies Global

## What is R-Ladies?

- A global community of ladies who use R
- that seeks to support and inspire others
- through teaching, sharing, and mentorship

# R-Ladies Bergen

## Who are we?

- Jonelle Dickow Villar
- Julia Romanowska
- Astrid Lunde

## Meetup group

- ca. 50 members already!
- who are we?
- why did we gather here?

# Poll results

## How advanced in R are you?

```{r poll1-data,echo=FALSE}
# here - create the tibble
# I want to learn!
# I’ve learned basics and I want more!
# R is one of many programming languages I know
# I am Hadley Wickham (she/her) ;)

poll.q1 <- tribble( ~q1.alt, ~answers,
                    "I want to learn!", NA,
                    "I’ve learned basics and I want more!", NA,
                    "R is one of many programming languages I know", NA,
                    "I am Hadley Wickham (she/her) ;)", NA
                  )
ggplot( poll.q1, aes( q1.alt, answers ) ) +
  geom_col()
```

# Aim of the meeting

## Aims

- Get to know one another
- Learn something new in R
- Get ideas for future meetings
