---
title: "R-Ladies Bergen Kick-off Meeting"
author: "Julia & Jonelle"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    logo: R-Ladies-horiz-bergen.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library( ggplot2 )
library( tidyverse )
```

# R-Ladies Global

## What is R-Ladies?

- A global community of ladies who use R
- that seeks to support and inspire others
- through teaching, sharing, and mentorship

# R-Ladies Bergen

## Who are we?

- Jonelle Dickow Villar
- Julia Romanowska
- Astrid Lunde

## Meetup group

- ca. 50 members already!
- who are we?
- why did we gather here?

# Poll results

## How advanced in R are you?

```{r poll1-data,echo=FALSE}
# updated by Julia on 22 Nov.2019

# here - create the tibble
poll.q1 <- tribble( ~q.alt, ~answers,
                    "I want to learn!", 2,
                    "I’ve learned basics and I want more!", 7,
                    "R is one of many programming languages I know", 3,
                    "I am Hadley Wickham (she/her) ;)", 1
                  )
# explicitly setting the order of labels
poll.q1$q.alt <- factor( poll.q1$q.alt, levels = rev( poll.q1$q.alt ), ordered = TRUE )

ggplot( poll.q1, aes( q.alt, answers, fill = q.alt ) ) +
  geom_col() + coord_flip() +
  theme_minimal() +
  theme( 
    plot.title = element_text( hjust = 1 ),
    plot.subtitle = element_text( hjust = 1 ),
    legend.position = "none"
  ) +
  ylab("") + xlab("") +
  labs( title = "Total of 13 R-Ladies answered", subtitle = "(out of 40 registered on Meetup)" )
```

## What is the field where you (would) use R the most?

```{r poll2-data,echo=FALSE}
poll.q2 <- tribble( ~q.alt, ~answers,
    "Bioinformatics / structural bioinformatics", 3,
    "Statistics", 6,
    "Applied mathematics", 0,
    "Political science", 0,
    "Economics", 1,
    "Visualization", 0,
    "History / Literature", 0,
    "Meteorology", 0,
    "Cheminformatics", 0,
    "Biology", 1,
    "AI / Machine learning", 0,
    "Other", 3
)
poll.q2$q.alt <- factor( poll.q2$q.alt, levels = rev( poll.q2$q.alt ), ordered = TRUE )

ggplot( poll.q2, aes( q.alt, answers, fill = q.alt ) ) +
  geom_col() + coord_flip() +
  theme_minimal() +
  theme( 
    plot.title = element_text( hjust = 1 ),
    plot.subtitle = element_text( hjust = 1 ),
    legend.position = "none"
  ) +
  ylab("") + xlab("") +
  labs( title = "Total of 13 R-Ladies answered", subtitle = "(out of 40 registered on Meetup)" )
```

## What do you use R for?

```{r poll3-data,echo=FALSE}
poll.q3 <- tribble( ~q.alt, ~answers,
    "Plotting", 0,
    "Organizing tables", 0,
    "Exploring the data", 1,
    "Specific analyses", 0,
    "Everything", 10,
    "Nothing (yet!)", 2
)
poll.q3$q.alt <- factor( poll.q3$q.alt, levels = rev( poll.q3$q.alt ), ordered = TRUE )

ggplot( poll.q3, aes( q.alt, answers, fill = q.alt ) ) +
  geom_col() + coord_flip() +
  theme_minimal() +
  scale_y_continuous( breaks = seq( 0, 10, 2 ), labels = as.character( seq( 0, 10, 2 ) ) ) +
  theme( 
    plot.title = element_text( hjust = 1 ),
    plot.subtitle = element_text( hjust = 1 ),
    legend.position = "none"
  ) +
  ylab("") + xlab("") +
  labs( title = "Total of 14 R-Ladies answered", subtitle = "(out of 40 registered on Meetup)" )
```

# Aim of the meeting

## Aims

- Get to know one another
- Learn something new in R
- Get ideas for future meetings
