---
title: "Plane crash data visualized"
author: "Julia Romanowska"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
---

```{r intro,include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library( tidyverse )
library( magrittr )
```

# Importing data

First, lets fetch the data, from years 2010 through 2019.

```{r read_data}
data <- read_delim( "plane_crash_2010-2019.dat", delim = "|" )
```

* tidy the data

```{r tidy_dt}
# pick relevant variables
data %<>% select( date, location, route, type, operator, aboard, fatalities, route )
data
```

# Explore the data 

We need to change the type of the `date` column

```{r date_type_change}
class( data$date )
head( data$date )

# change the locale temporary, so that the `as.Date` will recognize month names
old.locale <- Sys.getlocale( category = "LC_TIME" )
Sys.setlocale( category = "LC_TIME", locale = "en_GB.utf8" )

data$date <- as.Date( data$date, format = "%B %e, %Y" )
data
Sys.setlocale( category = "LC_TIME", locale = old.locale )
```

Extract year and month from each date.

```{r date_extract}
yr.month.date <- lapply( data$date, function(x){
  tribble( ~year, ~month,
           format( as.Date( x ), "%Y" ), format( as.Date( x ), "%m" ) )
})
yr.month.date %<>% bind_rows()
yr.month.date

data <- bind_cols( data, yr.month.date )
```

Are there any specific airlines that had more crashes?

```{r}
sort( table( data$operator ) )
```

The `aboard` column includes more info

```{r parse_aboard}
head( data$aboard )
aboard.list <- lapply( data$aboard, function(x){
  # extract the first number
  total <- str_extract( x, "^[[:digit:]]+" )
  passengers <- unlist( str_split( str_extract( x, "passengers:[[:digit:]]+"), ":" ) )[2]
  crew <- unlist( str_split( str_extract( x, "crew:[[:digit:]]+"), ":" ) )[2]
  return( tribble( ~aboard.total, ~aboard.pass, ~aboard.crew,
                   as.integer( total ) , as.integer( passengers ), as.integer( crew ) ) )
})
aboard.tbl <- bind_rows( aboard.list )
aboard.tbl

# merge this new information into our table
data <- bind_cols( select( data, -aboard ), aboard.tbl )
```

## Exercise 1

Parse the `fatalities` column in a similar way

```{r}

```

## Exercise 2

What are the levels in the `type` column?

```{r}

```


# Plot

Let's check whether the crashes have any pattern.

* total number of crashes per year

```{r group_by_year}
ggplot( data, aes( year ) ) +
  geom_bar()
```

* total number of passengers aboard all planes per year

```{r passengers_group_per_year}
ggplot( data, aes( aboard.pass ) ) +
  geom_bar() +
  facet_wrap( vars( year ), ncol = 5 )
```

## Exercise 3

Plot total number of crashes per month, per year

```{r}

```

